
\chapter{Preprocessing}

say something about the Preprocessing stage of compilation here.

\section{The define stack}

explain what the define stack is.


The function \verb|define_push()| is used to add to the define stack. It does this by first associating a lexeme to a symbol object, then associating this symbol to a token. This symbol object can only be one of two kinds of macro types \verb|MACRO_OBJ| or \verb|MACRO_FUNC|.


\begin{verbatim}
1317 ST_INLN void define_push(int v, int macro_type, int *str, Sym *first_arg)
1318 {
1319     Sym *s, *o;
1320
1321     o = define_find(v);
1322     s = sym_push2(&define_stack, v, macro_type, 0);
1323     s->d = str;
1324     s->next = first_arg;
1325     table_ident[v - TOK_IDENT]->sym_define = s;
1326 
1327     if (o && !macro_is_equal(o->d, s->d))
1328         tcc_warning("%s redefined", get_tok_str(v, NULL));
1329 }
\end{verbatim}
\begin{tcc_desc}
\textbf{v} \verb|v| is the token for the identifier.

\textbf{macro\_type} there are only two kinds of macros - \verb|MACRO_OBJ| which is used to describe object-like macros, and \verb|MACRO_FUNC| which describes function like macrosses. 

\textbf{str} the lexeme associated with this token.

\textbf{first\_arg} first\_arg param

\textbf{line 1325} the token to this was already allocated and added prior to the same index via \verb|tok_alloc_new()|

\end{tcc_desc}

\section{Finding a define}
\begin{verbatim}
ST_INLN Sym *define_find(int v)
{
    v -= TOK_IDENT;
    if ((unsigned)v >= (unsigned)(tok_ident - TOK_IDENT))
        return NULL;
    return table_ident[v]->sym_define;
}
\end{verbatim}


\section{Defining Object-Like Macros}

The tokenizer sees the '\#' character and determines that a preprocessing operation would be happening next. the tokenizer gets called for the second time to extract the next few characters - "define" in this case and a hash table
index is computed using the characters. the symbol entry for "define" is searched in the hash table and naturally, it will be found, so a new token will not be allocated, but instead, a reference to the symbol for "define" will be returned.

the global tok variable is set to the token id for "define" so it can be used by other parts of the compiler.

the tokenizer is called for the third time to extract the macro identifier. A new token symbol is created for this identifer and it is added to the hash table.

the global tok variable is set to the token id for this indentifier now.

now it is time to parse the object-like macro. the function parse\_define() is called to do this.

the tokenizer is called a fourth time to extract the value now. when tok is TOK\_PPNUM, the global variable tokc holds the token value and the global tok holds TOK\_PPNUM . TOK\_PPNUM is obviously not a space and the function tok\_str\_add2() to be called so that token id, length of the toke, and value of the token will be appended to the global token string tokstr\_buf.

tok\_str\_add2() utilizes the global tok value to do its operation and when it sees that it is looking at at TOK\_PPNUM token, it will consult the global CValue to extract the actual value.

A last call to the tokenizer will simply update the internal tracking of the tokenizer to point to next line.




% preprocessing-token: pp-number

% pp-number:
%       digit
%       . digit
%       pp-number digit
%       pp-number identifier-nondigit
%       pp-number e sign
%       pp-number E sign
%       pp-number p sign
%       pp-number P sign
%       pp-number .

% defining simple ints
% simple ints fall under the digit production rule 

% c => peeked character 
% t => former character
%
% t is already checked to be valid. (initially it is checked to be a number)
% now check if the peeked character is also valid.
% 
% (!(         (isidnum_table[c - CH_EOF] & (IS_ID|IS_NUM)) // check if peeked char is a number. => digit production rule
%          || c == '.'                                     // if peeked char is not a number, is it a decimal sign? => pp-number . production rule


%             // if peeked char is a sign '+' or '-', 
%             // then the former must be 'e' or 'E'       +e or +E => e sign or E sign production rule
%             // or 'p' or 'P'                            +p or +P => p sign or P sign production rule
%          || ( (c == '+' || c == '-') && (     (    (t == 'e' || t == 'E')
%                                                 && !(parse_flags & PARSE_FLAG_ASM_FILE && ((char*)tokcstr.data)[0] == '0' && toup(((char*)tokcstr.data)[1]) == 'X')
%                                               )
%                                            || t == 'p' || t == 'P'
%                                         )
%             )
%    )
% )

% the . digit production rule is handled by the case '.' ....


when the macro is just a simple int value:
        set the global variable to hold a constant value \verb|tokc|
        tokc.size = \# of digits
        tokc.data = simple int in string form
        tok = TOK\_PPNUM


\subsubsection{Transferring a Constant Value to a string}

The global variable \verb|tokc| holds a valid constant value when the global \verb|tok| value is \verb|TOK_PPNUM|. Outlined below are the specific lines of \verb|tok_str_add2()| function handling a \verb|TOK_PPNUM| token.

\begin{verbatim}
1145 static void tok_str_add2(TokenString *s, int t, CValue *cv)
1146 {
1147     int len, *str;
1148 
1149     len = s->lastlen = s->len;
1150     str = s->str;
1151 
1152     /* allocate space for worst case */
1153     if (len + TOK_MAX_SIZE >= s->allocated_len)
1154         str = tok_str_realloc(s, len + TOK_MAX_SIZE + 1);
1155     str[len++] = t;
         ...
1169     case TOK_PPNUM:
1170     case TOK_PPSTR:
1171     case TOK_STR:
1172     case TOK_LSTR:
1173         {
1174             /* Insert the string into the int array. */
1175             size_t nb_words =
1176                 1 + (cv->str.size + sizeof(int) - 1) / sizeof(int);
1177             if (len + nb_words >= s->allocated_len)
1178                 str = tok_str_realloc(s, len + nb_words + 1);
1179             str[len] = cv->str.size;
1180             memcpy(&str[len + 1], cv->str.data, cv->str.size);
1181             len += nb_words;
1182         }
1183         break;
1212     default:
1213         break;
1214     }
1215     s->len = len;
1216 }
\end{verbatim}

\begin{tcc_desc}
        \textbf{line 1149} s->len is counted in terms of words, not byte.

        \textbf{line 1149-1150} local references to TokenString object.

        \textbf{line 1153-1154} resize (enlarge) the TokenString object buffer so it can fit more data.

        \textbf{line 1155} field 0 when representing a PPNUM holds the token.

        \textbf{line 1169-1172} all of these 4 token types have the same handling.

        \textbf{line 1175-1178} \verb|tok_str_realloc()| allocates in terms of the system integer width (usually 4 bytes, also called a "word"). This line computes the number of words necessary to cover the string length. e.g. if string length is 3, then we only need 1 word. If however we have a string length of 5, then it rounds up the word count to 2. Additionally, +1 word is added to be able to save the string length itself into field 1.

        \textbf{line 1179} field 1 is used to store the string length.

        \textbf{line 1180} the actual string gets copied here into field 2.

        \textbf{line 1181 and line 1215} reflect the new length now.
\end{tcc_desc}

It is important to remember that TokenString keeps strings as an array of words (or ints) which is usually 4 bytes. Therefore, the length field inside a TokenString object is not the actual string length. Instead, it reflects the length of the buffer in terms of words.

\begin{verbatim}
1478 ST_FUNC void parse_define(void)
1479 {
1480     Sym *s, *first, **ps;
1481     int v, t, varg, is_vaargs, spc;
1482     int saved_parse_flags = parse_flags;
1483
1484     v = tok;
1485     if (v < TOK_IDENT || v == TOK_DEFINED)
1486         tcc_error("invalid macro name '%s'", get_tok_str(tok, &tokc));
1487     /* XXX: should check if same macro (ANSI) */
1488     first = NULL;
1489     t = MACRO_OBJ;
....    
1494     parse_flags = ((parse_flags & ~PARSE_FLAG_ASM_FILE) | PARSE_FLAG_SPACES);
....    
1496     next_nomacro_spc();
1497     if (tok == '(') {
....
1527     }
1528
1529     tokstr_buf.len = 0;
1530     spc = 2;
1531     parse_flags |= PARSE_FLAG_ACCEPT_STRAYS | PARSE_FLAG_SPACES | PARSE_FLAG_LINEFEED;
....
1537     while (tok != TOK_LINEFEED && tok != TOK_EOF) {
1538         /* remove spaces around ## and after '#' */
1539         if (TOK_TWOSHARPS == tok) {
1540             if (2 == spc)
1541                 goto bad_twosharp;
1542             if (1 == spc)
1543                 --tokstr_buf.len;
1544             spc = 3;
1545             tok = TOK_PPJOIN;
1546         } else if ('#' == tok) {
1547             spc = 4;
1548         } else if (check_space(tok, &spc)) {
1549             goto skip;
1550         }
1551         tok_str_add2(&tokstr_buf, tok, &tokc);
1552     skip:
1553         next_nomacro_spc();
1554     }
1555
1556     parse_flags = saved_parse_flags;
1557     if (spc == 1)
1558         --tokstr_buf.len; /* remove trailing space */
1559     tok_str_add(&tokstr_buf, 0);
1560     if (3 == spc)
1561 bad_twosharp:
1562         tcc_error("'##' cannot appear at either end of macro");
1563     define_push(v, t, tok_str_dup(&tokstr_buf), first);
1564 }
\end{verbatim}



\section{Defining Function-Like Macros}


\section{Finding a define Identifier}

The function \verb|define_find()| returns a reference to the symbol object associated with a token \verb|v|. It is outlined below with explanation.

\begin{verbatim}
ST_INLN Sym *define_find(int v)
{
    v -= TOK_IDENT;
    if ((unsigned)v >= (unsigned)(tok_ident - TOK_IDENT))
        return NULL;
    return table_ident[v]->sym_define;
}
\end{verbatim}

\begin{tcc_desc}
\textbf{line 244} normalize the token in reference to \verb|TOK_IDENT|.

\textbf{line 245-246} The global token counter \verb|tok_ident| is the source for all token numbers. Therefore, a valid token is always less than the global token counter (that is normalized in reference to \verb|TOK_IDENT|) which is what is checked here. Anything greater than, is obviously a non-existent token as far as TinyCC is concerned.

\textbf{line 247} return the define tokensymbol for token \verb|v|.
\end{tcc_desc}


\section{Registering a define macro to the system - The Define Stack}

The define stack where define symbols are pushed as they are discovered. The function to push new macrosses to the TinyCC define stack is \verb|define_push()|. It is outlined below and explained line by line.

\begin{verbatim}
1317 ST_INLN void define_push(int v, int macro_type, int *str, Sym *first_arg)
1318 {
1319     Sym *s, *o;
1320
1321     o = define_find(v);
1322     s = sym_push2(&define_stack, v, macro_type, 0);
1323     s->d = str;
1324     s->next = first_arg;
1325     table_ident[v - TOK_IDENT]->sym_define = s;
1326
1327     if (o && !macro_is_equal(o->d, s->d))
1328         tcc_warning("%s redefined", get_tok_str(v, NULL));
1329 }
\end{verbatim}

\begin{tcc_desc}
\textbf{line 1321} return a reference to this identifier - if has been defined prior.

\textbf{line 1322-1324} create and push a new define symbol now onto the define stack \verb|define_stack|. Initialize this symbol with additional information pertaining to the macro the compiler just parsed.

\textbf{line 1325} store a reference to this actual symbol now inside the token symbol table.

\textbf{line 1327-1328} detect if this particular macro identifier has already been prior. Then, alert user of the fact.

\end{tcc_desc}

\section{Undefining}

\section{Including Files}

\section{Error}

\section{Pragmas}

\section{Conditional Preprocessing}

\subsubsection{Ifdef}
\subsubsection{Ifndef}
\subsubsection{Elif}

\subsubsection{Ending a Conditional Block}


